---
title: "carnation - airway tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{carnation - airway tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries & airway dataset

Load carnation.

```{r setup}
library(carnation)
```

First load some libraries that we will need for this tutorial.

```{r message=FALSE}
library(DESeq2)
library(dplyr)
library(GeneTonic)

# install optional packages if not present
pkgs_to_check <- c('airway', 'org.Hs.eg.db', 'clusterProfiler', 'DEGreport')
for(pkg in pkgs_to_check){
  setRepositories(ind=c(1,2,3,4,5))
  if(!requireNamespace(pkg, quietly=TRUE)){
    install.packages(pkg, repos='http://cran.us.r-project.org')
  }
  library(pkg, character.only=TRUE)
}
```

We will be using the 'airway' dataset. First, we load this dataset.

```{r}
data('airway')
```

Next, we extract the counts matrix and and metadata.

```{r}
mat <- assay(airway)
cdata <- colData(airway)
```

Now let's see what these look like.

```{r}
dim(mat)
```

So, `mat` is a matrix with 64102 rows and 8 columns. Each row
corresponds to a single gene and each column corresponds to a
single sample. As you will notice, the rownames of `mat` contains
gene IDs and column names have the sample IDs.

```{r}
head(mat)
```

`cdata` contains the sample metadata. There is a lot of information here,
but notice the `cell` and `dex` columns, as we will be using this for the
differential expression analysis later.

```{r}
cdata
```

## Get more gene annotation

The gene IDs that come with the dataset are from ENSEMBL and are not human-readable.
So, next we will extract gene symbols and `ENTREZID` for these genes from
the `org.Hs.eg.db` package.

```{r message=FALSE}
keytypes <- list('SYMBOL'='SYMBOL', 'ENTREZID'='ENTREZID')

anno_df <- do.call('cbind',
             lapply(keytypes, function(x){
               mapIds(org.Hs.eg.db,
                 column=x,
                 keys=rownames(mat),
                 keytype='ENSEMBL')
               })
             )

# convert to data frame
anno_df <- as.data.frame(anno_df)
```

Now, we have human readable gene names in the `SYMBOL` column and Entrez IDs in the
`ENTREZ` column.

```{r}
head(anno_df)
```

## Create DESeqDataSet

Next, we create a new `DESeqDataSet` using `mat` and `cdata`.

```{r}
dds <- DESeqDataSetFromMatrix(mat,
                              colData=cdata,
                              design=~cell + dex)
```

Let's check to make sure that everything looks okay:

```{r}
dds
```

Then we save `dds` in a list.

```{r}
dds_list <- list(main=dds)
```

We also normalize the counts data and save it in a list.

```{r}
rld_list <- lapply(dds_list, function(x) varianceStabilizingTransformation(x, blind=TRUE))
```

## Run differential expression analysis

Now, the object `dds` is ready for differential expression (DE) analysis.

```{r}
dds <- DESeq(dds)
```

Since, we used `design=~cell + dex` while creating `dds`, the above step
will automatically calculate some comparisons.

```{r}
resultsNames(dds)
```

The last comparison `dex_untrt_vs_trt` contains the effect of the `dex` treatment,
while the other comparisons compare different cell lines. Next, we will extract
two of these results and run `lfcShrink` on them.

For the `cell` comparison, we choose a precomputed results using the `coef` parameter.

```{r message=FALSE}
cell_comparison <- lfcShrink(dds,
                             coef='cell_N61311_vs_N052611',
                             type='normal')
```

For the `dex` comparison, we use `contrast` to specify the direction of the comparison,
since we want to use `untrt` as control.

```{r message=FALSE}
dex_comparison <- lfcShrink(dds,
                            contrast=c('dex', 'trt', 'untrt'),
                            type='normal')
```

Now, each of these comparisons, contain the DE analysis results. For example,

```{r}
head(dex_comparison)
```

Then, we save these results in a special nested list that `carnation` will use.
Here,

- `res` contains the actual DE analysis results
- `dds` contains the name of the DESeqDataSet used for the DE analysis. These values
  should map to `dds_list` names
- `label` is a description of the comparison

```{r}
res_list <- list(
        dex_trt_vs_untrt=list(
            res=dex_comparison,
            dds='main',
            label='dex, treated vs untreated'),
        cell_N61311_vs_N052611=list(
            res=cell_comparison,
            dds='main',
            label='cell, N61311 vs N052611')
        )
```

Finally, we add `SYMBOL` and `ENTREZID` columns to the DE results from the `anno_df`
data frame.

```{r}
res_list <- lapply(res_list, function(x){
              # save the rownames as a new 'gene' column
              x$res[[ 'gene' ]] <- rownames(x$res)

              # add 'SYMBOL' and 'ENTREZID' columns
              x$res[[ 'SYMBOL' ]] <- anno_df[rownames(x$res), 'SYMBOL']
              x$res[[ 'ENTREZID' ]] <- anno_df[rownames(x$res), 'ENTREZID']

              x
            })
```

## Add functional enrichment results (optional)

Now we run functional enrichment on the DE genes from the two comparisons.
For this, we first set significance thresholds and then extract the DE genes
and save as a list.

```{r}
# padj cutoff
alpha <- 0.01

# log2FoldChange threshold; 1 == 2x difference
lfc_threshold <- 1

# list to save DE genes
de.genes <- lapply(res_list, function(x){
              # changed genes
              idx <- x$res$padj < alpha &
                     !is.na(x$res$padj) &
                     abs(x$res$log2FoldChange) >= lfc_threshold

              # return DE genes as a dataframe
              x$res[idx, c('gene', 'ENTREZID')]
            })
```

Next, we run functional enrichment and save the results in a list
called `enrich_list`. We also save a converted list called `genetonic`
which carnation uses for several plots from the `GeneTonic` package.

```{r}
# list to save functional enrichment results
enrich_list <- list()

# list to save a converted object for GeneTonic plots
genetonic <- list()

for(comp in names(res_list)){
    go.res <- clusterProfiler::enrichGO(
            gene=de.genes[[comp]][['ENTREZID']],
            keyType='ENTREZID',
            OrgDb=org.Hs.eg.db,
            ont='BP',
            pvalueCutoff=1, qvalueCutoff=1,
            readable = TRUE)

    enrich_list[[ comp ]] <- list(
                               res=comp,
                               changed=list( BP=as.data.frame(go.res) )
                             )

    genetonic[[ comp ]] <- list(
                             res=comp,
                             changed=list(
                               BP=enrich_to_genetonic(go.res, res_list[[comp]]$res)
                             )
                           )

}
```

`enrich_list` is a nested list where:

- The top-level names are unique keys/identifiers
- The second level corresponds to direction of change, e.g. `up`, `down` or `changed`.
  This level also contains a special entry `res` which maps to `res_list` names,
  as a way to record where the DE results came from.
- The third level corresponds to the ontology, e.g. `BP` (GO Biological Process).

Here, we are just using changed genes and the GO Biological Process (BP) ontology,
so `enrich_list` looks like:

```
enrich_list
  ├─ dex_trt_vs_untrt
  │    ├─ res = 'des_trt_vs_untrt'    <--- comparison used to get FE results
  │    └─ changed
  │         └─ BP                     <--- functional enrichment results
  │
  └─ cell_N61311_vs_N052611
       ├─ res = 'cell_N61311_vs_N052611'    <--- comparison used to get FE results
       └─ changed
            └─ BP                           <--- functional enrichment results

```

`genetonic` mirrors the same structure.

## Add pattern analysis (optional)

Finally, we add some pattern analysis for the `dex_trt_vs_untrt` comparison using
the `DEGreport` package. First, we extract normalized data for the 755 DE genes from
this comparison.

```{r}
# extract normalized data & metadata
ma <- assay(rld_list[['main']])
colData.i <- colData(rld_list[['main']])

# only keep data from DE genes
idx <- rownames(ma) %in% de.genes[['dex_trt_vs_untrt']][['gene']]
ma.i <- ma[idx,]

# remove any genes with 0 variance
ma.i <- ma.i[rowVars(ma.i) != 0, ]
```

Then, we run the pattern analysis, using `cell` as the *time*
variable and `dex` as the *color* variable.

```{r}
p <- DEGreport::degPatterns(
        ma.i,
        colData.i,

        time='cell',
        col='dex',

        # NOTE: reduce and merge cutoff----------------------------------------
        #   Reduce will merge clusters that are similar; similarity determined
        #   by cutoff
        reduce=TRUE,

        # NOTE: set minimum cluster size to 1
        minc=1,
        plot=FALSE
        )
```

Next, we extract the `normalized` slot from this object and save as a list.

```{r}
# extract normalized slot and add symbol column
p_norm <- p$normalized
p_norm[[ 'SYMBOL' ]] <- anno_df[p_norm[['genes']], 'SYMBOL']

# save pattern analysis results
degpatterns <- list(dex_by_cell=p_norm)
```

## Compose carnation object

Now we have all the pieces to build the carnation object.

```{r}
combined <- list(res_list=res_list,
                 dds_list=dds_list,
                 rld_list=rld_list,
                 enrich_list=enrich_list,
                 genetonic=genetonic,
                 degpatterns_list=degpatterns)
saveRDS(combined, 'carnation_vignette.rds', compress=FALSE)
```

# sessionInfo

```{r}
sessionInfo()
```
